Supplemental materials for ``An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)''

### 2022-05-12 manuscript/main.tex
updated clone residual variance args + updated supp tables

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of item template (level 1) attributes (described below) on difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

\item $\epsilon_{\beta j}$: the template (level 1) residual, or the residual variability in difficulty of the item template unexplained by its attributes.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of item clone (level 2) attributes on difficulty, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$.

\item $\epsilon_{\beta k}$: the clone (level 2) residual, or residual variability in difficulty of the item clone unexplained by its attributes.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. By comparison, Model 5 demonstrated smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 6 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that Model 6 fit clearly better than Model 4 ($\Delta$ LOCO = 85.91, se = 6.33, supporting the inclusion of varying item discrimination parameters). We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

\subsubsection{Aim 1: Estimate item parameters for all items in the MaRs-IB}

Across all items, the average item difficulty was $\mu_\beta = 0.177$ (95\% HDI = 0.118 -- 0.238). There was considerable variability in item difficulty across items (sd = 1.431, 95\% HDI = 1.370 -- 1.492), wih the smallest ($\beta$ = -3.704, 95\% HDI = -4.687 -- -2.775) and largest ($\beta$ = 3.497, 95\% HDI = 2.481 -- 4.635) item difficulty parameters spanning a wide range. For an average ability participant ($\theta = 0$), these parameter estimates translate to an expected average accuracy of 58.4\% across all items, and ceiling (floor) performance for the least (most) difficult item.

@@

\subsubsection{Aim 2:}

%\subsubsection{Aim 2: Quantify the associations between item attributes and functioning}

@@

%\subsubsection{Objective 3: Decompose the variability in item functioning}

@@

The estimated item difficulty per clone, organized by distractor type and shape set, is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast, the pattern of item difficulties by shape set is more complicated (Figure \ref{fig:fig02}b). We computed post-hoc contrasts to compare item difficulty within items across shape sets. We found that item clones of shape set 2 were marginally more difficult on average than clones of shape set 1 ($\bar{\beta}_{ss_2} - \bar{\beta}_{ss_1}$ = 0.149, 95\% HDI = 0.054 -- 0.241); and clones of shape set 3 were marginally more difficult on average than clones of shape set 2 ($\bar{\beta}_{ss_3} - \bar{\beta}_{ss_2}$ = 0.161, 95\% HDI = 0.066 -- 0.259). However, the magnitude of these differences were still a fraction of the magnitude of the clone-level residual variance ($\epsilon_{\beta k}$ =  0.620, 95\% HDI = 0.538 -- 0.701). Thus, the sources of residual variability in item difficulty across clones is complex and likely idiosyncratic based on  perceptual features...

@@

We calculated the ordinal reliability of the MaRs-IB and RPM short-form measures using the lavaan \citep{lavaan} and semTools \citep{semtools} packages available in R.

@@

\begin{table}
    \centering
    \begin{tabular*}{1.02\textwidth}{cccccccccccr}
    \toprule
    && \multicolumn{3}{c}{Difficulty} && \multicolumn{3}{c}{Discrimination} && \multicolumn{2}{c}{LOO-CV} \\
    \cmidrule(lr){3-5}\cmidrule(lr){7-9}\cmidrule(lr){11-12}
    Model && {\small FE-1/2} & {\small RE-1} & {\small RE-2} && {\small FE-1/2} & {\small RE-1} & {\small RE-2} && psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    1 && X &   &   &&   &   &   && 27741.4 & 1326.87 (62.88) \\
    2 && X & X &   &&   &   &   && 26962.1 & 547.56 (37.79) \\
    3 && X & X & X &&   &   &   && 26585.5 & 171.05 (12.65) \\
    4 && X & X & X && X &   &   && 26419.6 & 5.12 (4.07) \\
    5 && X & X & X && X & X &   && 26414.5 & \multicolumn{1}{c}{-} \\
    6 && X & X & X && X & X & X && 26417.8 & 3.29 (1.84) \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Goodness-of-fit of item response models to the MaRs-IB response data. The columns under difficulty and discrimination indicate the specification of those parameters for each model (i.e. the presence of level-1/2 fixed effects, level-1 random effects, and level-2 random effects). LOO-CV values are presented in deviance scale (i.e. smaller values indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}

>>>>

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of item template (level 1) attributes (described below) on difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$. This component is the fixed effects contribution to item template difficulty.

\item $\epsilon_{\beta j}$: the template (level 1) residual, or the residual variability in difficulty of the item template unexplained by its attributes. This component is the random effects contribution to item template difficulty.

\item $\sum_{m=1}^M R_{km} \delta_{\beta m}$: the effect of item clone (level 2) attributes on difficulty, where $R_{km}$ is the value of attribute $m$ for item clone $k$, and $\delta_{\beta m}$ is the effect of attribute $m$. This component is the fixed effects contribution to item clone difficulty.

\item $\epsilon_{\beta k}$: the clone (level 2) residual, or residual variability in difficulty of the item clone unexplained by its attributes. This component is the random effects contribution to item clone difficulty.

@@

In the second wave of model fitting, in which models varied only in their specification of item discrimination, Model 5 demonstrated the best fit to the data. By comparison, Model 5 demonstrated smaller improvements in fit over Model 4 ($\Delta$ LOCO = 5.12, se = 4.07) and Model 6 ($\Delta$ LOCO = 3.29, se = 1.84). This result indicates the presence of non-negligible residual variance in item discrimination across item templates but not clones. It must be noted these differences in LOCO values from each model to the next were each within two standard errors of the mean, indicating only weak predictive improvement of Model 6 over the others \citep{vehtari2022cv}. As such we will select Model 5 as the best-fitting model overall, but proceed in discussing it with caution. We also note that Model 6 fit clearly better than Model 4 ($\Delta$ LOCO = 85.91, se = 6.33, supporting the inclusion of varying item discrimination parameters). We also note that all second-wave models exhibited better fits to the data than the first-wave models, supporting the inclusion of item discrimination parameters.

\subsubsection{Aim 1: Quantify the psychometric properties of items in the MaRs-IB}

Across all items, the average item difficulty was $\mu_\beta = 0.177$ (95\% HDI = 0.118 -- 0.238). There was considerable variability in item difficulty across items (sd = 1.431, 95\% HDI = 1.370 -- 1.492), with the smallest  and largest item difficulty parameters spanning a wide range ($\beta_{\min}$ = -3.704, 95\% HDI = -4.687 -- -2.775; $\beta_{\max}$ = 3.497, 95\% HDI = 2.481 -- 4.635). For an average ability participant ($\theta = 0$), these parameter estimates translate to an expected average accuracy of 58.4\% across all items, and ceiling (chance-level) performance for the least (most) difficult item.

@@

\subsubsection{Aim 2: Measuring the association between item complexity and functioning}

@@

\subsubsection{Objective 3: Investigating the exchangeability of item clones}

@@

The estimated item difficulty per clone, organized by distractor type and shape set, is presented in Figure \ref{fig:fig02}. The relative increase in difficulty for items with MD-type distractors compared to PD-type distractors is visually apparent (Figure \ref{fig:fig02}a). In contrast, the pattern of item difficulties across clones by shape set is more complicated (Figure \ref{fig:fig02}b). After controlling for distractor type and mean response time, the magnitude of the clone-level residual variance is $\epsilon_{\beta k}$ = 0.620 (95\% HDI = 0.538 -- 0.701). Across the 128 unique sets of three clones, organized by item and distractor type, that level of residual variance translates into an average difference in difficulty of 0.752 (95\% HDI = 0.630 -- 0.877) between the easiest and hardest item clones of each triad, an effect that is two-thirds as large as the effect of distractor type. Furthermore, the shape set membership of the most difficult clone varies considerably across the 128 sets of clones (shape set 1 = 39; shape set 2 = 32; shape set 3 = 57). Thus, the sources of residual variability in item difficulty across clones is complex and likely reflect the idiosyncratic contributions of their low-level perceptual features.

@@

We calculated the ordinal reliability of the MaRs-IB and RPM short-form measures using the \textit{lavaan} \citep{lavaan} and \textit{semTools} \citep{semtools} packages available in R.

@@

\begin{table}
    \centering
    \begin{tabular*}{1.02\textwidth}{cccccccccccr}
    \toprule
    && \multicolumn{3}{c}{Difficulty} && \multicolumn{3}{c}{Discrimination} && \multicolumn{2}{c}{LOO-CV} \\
    \cmidrule(lr){3-5}\cmidrule(lr){7-9}\cmidrule(lr){11-12}
    Model && {\small FE-1/2} & {\small RE-1} & {\small RE-2} && {\small FE-1/2} & {\small RE-1} & {\small RE-2} && psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    1 && X &   &   &&   &   &   && 27741.4 & 1326.87 (62.88) \\
    2 && X & X &   &&   &   &   && 26962.1 & 547.56 (37.79) \\
    3 && X & X & X &&   &   &   && 26585.5 & 171.05 (12.65) \\
    4 && X & X & X && X &   &   && 26419.6 & 5.12 (4.07) \\
    5 && X & X & X && X & X &   && 26414.5 & \multicolumn{1}{c}{-} \\
    6 && X & X & X && X & X & X && 26417.8 & 3.29 (1.84) \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Comparison of item response models fit to the MaRs-IB response data. The columns under difficulty and discrimination indicate the specification of those parameters for each model (i.e. the presence of level-1/2 fixed effects, level-1 random effects, and level-2 random effects). LOO-CV values are presented in deviance scale (i.e. smaller values indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}

### 2022-05-12 manuscript/supplement.tex
manuscript edits
updated clone residual variance args + updated supp tables

\begin{table}
    \centering
    \begin{tabular*}{\textwidth}{clcr}
    \toprule
    Model & Description & psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    7 & TK  & 27741.4 & 635.3 (29.7) \\
    8 & TK  & 26962.1 & 665.6 (32.2) \\
    9 & TK & 26585.5 & 633.1 (33.9) \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Goodness-of-fit of item response models to MaRs-IB data. LOCO values are presented in deviance scale (i.e. smaller numbers indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}

>>>>

\begin{table}
    \centering
    \begin{tabular*}{0.45\textwidth}{ccr}
    \toprule
    Model & psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    1r & 28376.7 & 635.3 (29.7) \\
    1 & 27741.4 & \multicolumn{1}{c}{-} \\
    \midrule
    2r & 27627.6 & 665.6 (32.2) \\
    2 & 26962.1 & \multicolumn{1}{c}{-} \\
    \midrule
    3r & 27218.7 & 633.1 (33.9) \\
    3 & 26585.5 & \multicolumn{1}{c}{-} \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Comparison of the first wave of item response models to their equivalents without a guessing parameter ($\gamma = 0$). LOO-CV values are presented in deviance scale (i.e. smaller values indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}
